Factors affecting inferences on natural mortality and associated environmental effects in state-space age-structured assessment models

### 2025-05-21 Ecov_study/mortality/manuscript/manuscript.Rmd
saving work on Ecov-M

State-space stock assessment models, with non-linear functions of latent processes and numerous observation types with different probability distribution assumptions represent one of most complex classes of state-space models. The literature on the ways we make inferences and the effects of various factors on reliabilty of inferences from state-space assessment models is growing [@lietal24;@milleretal_inreview1;@cadiganetal]. The importance of contrast in population size and fishing mortality and quality of data used to fit assessment models including the state-space variety is known [@magnussonhilborn07;@milleretal_inreview1]. Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios [@leeetal11;johnsonetal15;@cadigan16;@millerhyun18;@milleretal_inreview1].

The effects of temporal variation in recruitment via unspecified or specified environmental factors have been extensively investigated in both traditional assessment models and state space models [@myers98;@milleretal16; @haltuchpapers]. Reliability of estimating environmental and spawning biomass effects on recruitment requires a combination of strong effects, good age composition data quality, contrast in the environmental covariate and lower recruitment variability [@brittenetal_inreview;@milleretal_inreview1].

Temporal and environmental effects on growth and weight at age are also more important for short term projections and the reliability of estimation of those effects @correaetal23; @correaetalinreview.

@@

See @millerretal_inreview1 fro relevance of project 0 results (estimability of natural mortality). Some difficulty in distinquishing variation in natural mortality or effects of explicit covariates when variation in M random effects or Ecov is low relative to observation error.

@@

 Incorrect treatment of population attributes as temporally varying [@trijouletetal20; @liljestrandetal24] could lead to misidentification of stock status and biased population estimates, ultimately impacting fisheries management decisions [@legaultpalmer16; @szuwalskietal18; @croninpunt21]. Furthermore, biological, fishery, and observational processes are often confounded in catch-at-age data, which may adversely affect ability to distinguish between true process variability and observational error [@puntetal14; @stewartmonnahan17; @croninpunt21; @fischetal23; @lietalinreview_a].

@@

Estimation of natural mortality is known to be challenging in stock assessment models (Lee et al., others). However, estimation has been shown to be reliable in some situations @millerhyun18)

@@

Here we conduct a simulation study with operating models varying by degree of observation error uncertainty, sources of process error (M or survival), fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality. The simulations from these operating models are fitted with estimating models that make alternative assumptions for sources of process error, and whether (mean) M is estimated. We evaluate effects of these factors on convergence of fitted models, whether AIC can correctly determine the correct source of process error and correct assumption about covariate effects on natural mortality, and the degree of bias in relevant parameters and outputs of the assessment model.

@@

All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20;@stockmiller21;@milleretal_inreview1]. We used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3) for this simulation study. This package has been used extensively to configure OMs and EMs for several other simulation studies [@legaultetal23; @lietal24, @lietalinreview] and is used to assess many commercially important stocks in the Northeast US.

-------------- need to edit this part

There are many similarities to the configuration of OMs and EMs in simulation studies conducted by @milletetelinreview. We completed a simulation study with a number of OMs that can be categorized based on where process error random effects are assumed: abundance at age (R, R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q). For each OM assumption about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.

In total, we configured 72 OMs with alternative assumptions about the source and variability of process errors, level of observation error in indices and age composition data, and contrast in fishing pressure over time. We fitted 20 EMs to observations from each of 100 simulations where process errors were also simulated. For R+M, R+Sel, and R+q OMs, we used unique seeds for each simulation, but we inadvertently used the same 100 seeds for all R and R+S OMs. Each of the EMs made alternative assumptions about the source of process errors and whether natural mortality (or the median for models with process error in natural mortality) was estimated and whether a Beverton-Holt stock recruit relationship was estimated within the EM. Details of each of the operating and EMs are described below.

We did not use the log-normal bias-correction feature for process errors or observations described by @stockmiller21 for operating and EMs [@lietalinreview]. Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster. All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/ecov_study/mortality/code.

--------------

--------------------------- From project_0 paper

We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs in our simulation study [@millerstock20;@stockmiller21]. WHAM is an R package freely available via a github repository and is built on the Template Model Builder package [@kristensenetal16]. For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3). WHAM has also been used to configure OMs and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is currently used or accepted for use in management of numerous NEUS fish stocks [e.g., @nefsc22;@nefsc22a;@nefsc24].

We completed a simulation study with a number of OMs that can be categorized based on where process error random effects were assumed: recruitment (R, assumed present in all models), apparent survival (denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q). We refer to the (R+S) OMs as modeling apparent survival because on logscale the random effects ($\epsilon_{a,y}$) are additive to the total mortality (F+M) between numbers at age, thus they modify the survival term. However, as @stockmiller21 note, these random effects can be due to events other than mortality, such as immigration, emigration, missreported catch, and other sources of misspecification.  For each OM, assumptions about the magnitude of the variance of process errors and observations are required and the values we used were based on a review of the range of estimates from Northeast Unite States (NEUS) assessments using WHAM.

In total, we configured 72 OMs with alternative assumptions about the source and magnitude of process errors, magnitude of observation error in indices and age composition data, and contrast in fishing pressure over time. We fitted 20 EMs to observations generated from each of 100 simulations where process errors were also simulated. Each EM differed in assumptions about the source of process errors, whether natural mortality (or the median for models with process error in natural mortality) was estimated, and whether a Beverton-Holt SRR was estimated within the EM. Details of each of the OMs and EMs are described below.

We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for OMs and EMs to simplify interpretation of the study results [@lietalinreview]. All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code.

-------------------------

We used the Woods Hole Assessment Model (WHAM) to configure operating and estimation models in our simulation study [@millerstock20;@stockmiller21]. WHAM is an R package freely available as a github repository. For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3). This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, and black sea bass in the Northeast US.

We completed a simulation study with a number of operating models. Many of the characteristics of the operating models are the same as those used in the working paper by Miller et al. (WP 4). The factors defining the configuration of each operating model which are described in detail in subsequent sections include 
\begin{itemize}
\item two configurations of fishing histories, 
\item two configurations of index and age composition observation error, 
\item four configurations of latent environmental covariate process errors, 
\item two configurations of environmental covariate observation uncertainty, 
\item three levels of covariate effects on natural mortality, and 
\item three configurations of sources of process errors for temporal changes in abundances at age. 
\end{itemize} 
The full factorial of the levels of each factor results in 288 different operating models. We simulated 100 data sets for each operating model that included simulations of process errors.

For each simulated data set we fit a set of 12 estimating models. The factors defining the configuration or each estimating model which are also described in detail below include
\begin{itemize}
\item three configurations of sources of process errors for temporal changes in abundances at age, 
\item whether the (mean) natural mortality rate was estimated or assumed known at the true value, and
\item whether the effect of the environmental covariate on natural mortality was estimated.
\end{itemize}

We did not use the bias-correction feature for process errors or observations described by [@stockmiller21] for operating and estimating models. Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster. Code for completing the simulations and summarizing results can be found at https://github.com/timjmiller/SSRTWG/ecov_study/mortality.

@@

@@

The population model tracks 10 age classes: ages 1 to 10+ and we assume spawning occurs 1/4 of the way through the year. The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 and assumed known in all estimation models (Figure \ref{om_maturity}).

Weight at age was generated with a LVB growth function
$$
L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
$$
where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a L-W relationship such that 
$$
W_a = \theta_1 L_a^{\theta_2}
$$
where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$ (Figure \ref{om_waa}).

We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters (i.e., $\alpha$ and $\beta$) for all operating models. All post-recruit productivity components are constant in the NAA and survey selectivity process error operating models. Therefore steepness and unfished recruitment are also constant over the time period for those operating models [@millerbrooks21]. We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = F_{40\%} = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
\[
N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
\]
Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without process errors on natural mortality we assumed the rate was assumed 0.2. For OMs with process errors on natural mortality the mean rate was 0.2.

Two alternative fishing histories were used for operating models. In the first scenario, the stock experiences overfishing for the first 20 years and fishing at \Fmsy for the last 20 years. In the second scenario, the stock is fished at \Fmsy for the entire time period. The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from @wiedenmannetal19. @legaultetal23 also used similar approaches to defining fishing mortality histories for operating models.

Initial population was configured at the equilibrium distribution fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$ for the two alternative fishing histories. That is for a deterministic model, the age composition would not change over time when the fishing mortality was constant at the respective level.

For operating models with time-varying random effects or covariate effects on M, steepness is not constant, but we used the same alpha and beta parameters as other operating models this equates to a steepness and $R_0$ at the mean of the time series process for M. For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.

In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate. The latent covariate is assumed to be Gaussian AR1
$$
E_y|E_{y-1} \sim \text{N}\left(\mu_\text{Ecov}\left(1-\rho_\text{Ecov}\right) + \rho_\text{Ecov} E_{y-1}, \left(1-\rho_\text{Ecov}^2\right)\sigma^2_\text{Ecov}\right)
$$
with marginal mean $\mu_\text{Ecov}=0$ and variance $\sigma^2_\text{Ecov}$. The configuration of the latent environmental covariate in the operating models have one of two alternate assumptions about the marginal standard deviation and autocorrelation parameter:
\begin{itemize}
\item marginal standard deviation $\sigma_\text{Ecov} \in \{0.1, 0.5\}$
\item AR1 correlation $\rho_\text{Ecov} \in \{0, 0.5\}$
\end{itemize}

The observations of the latent environmental covariate are assumed to be unbiased and Gaussian
$$
\text{ecov}_y|E_y \sim \text{N}\left(E_y,\sigma^2_\text{ecov}\right)
$$
The standard deviation of the environmental observations in the operating models is one of two values:
\begin{itemize}
\item standard deviation $\in \{0.1, 0.5\}$
\end{itemize}
Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.

### Random effects on recruitment, survival, and natural mortality {-}

Internally, WHAM treats natural mortality as a log-transformed parameter 
$$
\log M_y = \beta_M + \beta_{\text{Ecov}} E_y + \varepsilon_{M,y}
$$
that is a linear combination of a mean $\beta_M$ and any annual covariate $\beta_{\text{Ecov}}$ and random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$ [@stockmiller21]. For operating models with natural mortality random effects, we assume the same standard deviation $\sigma_M = 0.3$. This is similar to the intermediate value used for R+M OMs by @milleretal_inreview1.

The covariate effect is one of 3 alternative values in the operating models:
\begin{itemize}
\item $\beta_\text{Ecov} \in \{0,0.25,0.5\}$
\end{itemize}
The parameters defining the simulated covariate time series, size of the covariate effect, and any natural mortality random effects result in a range of different levels of variation in annual natural mortality rates (Figure \ref{M_example}).

We configured all OMs with uncorrelated random effects on recruitment with standard devaition on log(recruitment) $\sigma_R = 0.5$ for R, R+S, and R+M OMs. This same assumption was used by @milleretalinreview for R+M OMs and other OMs with fishery selectivity and index catchability process errors. For R+S OMs, apparent survival process errors were uncorrelated  with $\sigma_{2+} = 0.3$ and for R+M OMs, process errors on natural mortality were uncorrelated with $\sigma_M = 0.3$.

### Fleets {-}

We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_mean_selectivity}). This selectivity was used to define \Fmsy, \Fspr[40], and steepness for the Beverton-Holt stock recruitment parameters above. We assumed a logistic-normal distribution for the age-composition observations associated with the fleet where errors in the multivariate normal transformation are independent. The standard deviation parameter is also constant across ages.

### Indices {-}

@@

### Observation Uncertainty {-}

@@

\begin{itemize}
\item whether the mean natural mortality $\beta_M$ was estimated or assumed known ($\log 0.2$)
\item whether an environmental effect $\beta_\text{Ecov}$ was estimated or not (fixed at 0)
\item whether the process errors were just on recruitment (R), on recruitment and survival (R+S) or on recruitment and natural mortality (R+M)
\end{itemize}
The configuration of the process errors in the estimating models generally matched the corresponding options in the operating models.

@@

## Measures of reliability {-}

@@

We calculated median errors of
\begin{itemize}
\item $\beta_\text{Ecov}$, the effect of environmental covariates on natural mortality, 
\item $\beta_M$, the mean log-natural mortality rate,
\end{itemize}
and the median relative errors of 
\begin{itemize}
\item annual natural mortality rate,
\item annual spawning stock biomass, and
\item annual fully-selected fishing mortality rate.
\end{itemize}

@@

For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the first year (start), year 21 (middel), and the last year (end), because there were no appreciable differences between the results for these three years and those for the other years.

@@

## Convergence performance {-}

When the (mean) log-natural mortality rate was assumed at the true value, corresponding to usual practice in application of assessment models, good convergence for all EMs was observed for R+S operating models with \Fmsy fishing history, more variation in the latent environmental covariate ($\sigma_\text{Ecov} = 0.5$), and lower error in the associated observations $\sigma_\text{ecov} = 0.1$ (Figure \ref{convergence_M_fixed}). Good convergence for all EMs was also observed for R+S OMs with the step change in fishing history, but with low error in indices and age comp observations.

EMs that assumed random effects just on recruitment (R), nearly always converged across all operating models that assumed more variation in the latent environmental covariate ($\sigma_\text{Ecov} = 0.5$) and lower error in the associated observations $\sigma_\text{ecov} = 0.1$. EMs that assumed recruitment and survival random effects (R+S), had poor converge probability when the OMs had alternative process errors (R or R+M). The R+S EMs showed best convergence for the R+S OMs where there was lower error in the environmental observations $\sigma_\text{ecov} = 0.1$ or higher error in environmental observations of latent true environmental covariates with greater temporal variation. EMs that assumed recruitment and natural mortality random effects (R+M), had poor convergence probability for all OMs that had a step change in fishing history and higher uncertainty in indices and age composition.

@@

As might be expected, there was an overall drop in the probability of convergence when the mean natural mortality rate was estimated rather than assumed at the true value (Figure \ref{convergence_M_estimated}). Otherwise, general trends described above with mean natural mortality fixed, apply when estimated.

@@

When estimating models assumed the mean natural mortality rate was known, the best accuracy of AIC for model selection occurred for models with R+S process errors. R+S estimating models ranked best with very low frequency for R or R+M operating models and with very high frequency for R+S operating models (Figure \ref{aic_rank_M_fixed}). R estimating models were determined best with high frequency for R operating models, but also for R+M operating models. R+M estimating models were rarely determined best for any operating models including those where the process errors matched.

@@

Relative to the assumption that the mean natural mortality rate was known, estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error and whether the covariate affect natural mortality (Figure \ref{aic_rank_M_estimated}). Where there were differences there were small decreases in accuracy for determining the appropriate process error and determining a covariate effect when there was one.

@@

When the EMs assumed the mean natural mortality rate was known, we observed generally accurate estimation of environmental effects across all EM and OM process error assumptions and all true covariate effect sizes, when ther was low uncertainty in environmental observations and larger temporal contrast in the simulated true environmental covariate (Figure \ref{Ecov_beta_bias_M_fixed}). We observed a negative trend in bias of the environmental effect with increased effect size when temporal variation in the covariate was lower and/or uncertainty in the covariate observations was higher. When the OMs had R+S process errors with low temporal variation in the true environmental covariate and lower uncertainty in the indices age age composition, estimated covariate effects were highly variable.  In most cases the relative error of $\beta_\text{Ecov}$ did not depend on the source of process error assumed in the EM. When there was an effect of the EM process error assumption it was when OMs had R+S process errors. The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.

@@

We present results for error in annual natural mortality rate conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assumed ($\beta_\text{Ecov} = 0$),  2) $\beta_M = log(0.2)$ and $\beta_\text{Ecov}$ is estimated, and 3) both $\beta_M$ and $\beta_\text{Ecov}$ are estimated.

When OMs and EMs assume $\beta_M = log(0.2)$ and $\beta_\text{Ecov} = 0$, there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{annual_M_bias_M_fixed_beta_fixed}). We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality ($\beta_\text{Ecov} > 0$). Including process error on M (R+M) produces more variability in errors for R and R+S EMs than including increased level of effect of the covariate on M. For R and R+S OMs, errors in annual M for R+S EMs were less variable than those for R EMs.

@@

>>>>

State-space stock assessment models, with non-linear functions of latent processes and numerous observation types with different probability distribution assumptions represent one of most complex classes of state-space models. The literature on the ways we make inferences and the effects of various factors on reliabilty of inferences from state-space assessment models is growing [@lietal24;@milleretal_inreview1;cadiganetal]. The importance of contrast in population size and fishing mortality and quality of data used to fit assessment models including the state-space variety is known [@magnussonhilborn07;@milleretal_inreview1]. Furthermore, estimation of natural mortality, and even temporal variablity is possible in many scenarios [@leeetal11;@johnsonetal16;@cadigan16;@millerhyun18;@milleretal_inreview1].

The effects of temporal variation in recruitment via unspecified or specified environmental factors have been extensively investigated in both traditional assessment models and state space models [@myers98;@milleretal16, haltuchpapers]. Reliability of estimating environmental and spawning biomass effects on recruitment requires a combination of strong effects, good age composition data quality, contrast in the environmental covariate and lower recruitment variability [@brittenetal_inreview;@milleretal_inreview1].

Temporal and environmental effects on growth and weight at age are also more important for short term projections and the reliability of estimation of those effects [@correaetal23; @correaetal_inreview].

@@

See @milleretal_inreview1 for relevance of project 0 results (estimability of natural mortality). Some difficulty in distinguishing variation in natural mortality or effects of explicit covariates when variation in M random effects or Ecov is low relative to observation error.

@@

 Mis-specified temporal population process errors  could lead to misidentification of stock status and biased population estimates and poor fisheries management decisions [@trijouletetal20; @legaultpalmer16; @szuwalskietal18; @croninpunt21;@liljestrandetal24].

@@

Estimation of natural mortality is known to be challenging in stock assessment models (@leeetal11, others). However, estimation has been shown to be reliable in some situations [@millerhyun18]

@@

Here we conduct a simulation study with OMs varying by degree of observation error uncertainty, sources of process error, fishing history, temporal variation in environmental covariates, and magnitude of the effect of the covariate on natural mortality. The simulations from these operating models are fitted with estimating models that make alternative assumptions for sources of process error, and whether (mean) M is estimated. We evaluate effects of these factors on convergence of fitted models, whether AIC can correctly determine the correct source of process error and correct assumption about covariate effects on natural mortality, and the degree of bias in relevant parameters and outputs of the assessment model.

@@

All of our analyses used the Woods Hole Assessment Model (WHAM) to construct both OMs and EMs [@millerstock20; @stockmiller21; @milleretal_inreview]. The WHAM package has been used extensively to configure OMs and EMs for several other simulation studies [@legaultetal23; @lietal24; @brittenetal_inreview; @lietal_inreview_a] and is used to assess many commercially important stocks in the Northeast U.S. [e.g., @nefsc22;@nefsc22a;@nefsc24]. We used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3) for to generate all results.

We completed a simulation study with a 288 operating models. The factors defining the configuration of each operating model which are described in detail in subsequent sections include source of population process error (3 levels), index and catch observation uncertainty (2 levels), environmental covariate uncertainty (2 levels), latent environmental covariate process stochasticity (4 levels), and fishing history (2 levels). We simulated 100 data sets for each operating model that included simulations of process errors.

For each simulated data set we fit a set of 12 estimating models (EMs). The factors that distinguish the estimating models which are also described in detail below include source of population process error type (3 levels) whether (median) natural mortality rate was estimated or assumed known (2 levels), and whether the effect of the environmental covariate on natural mortality was estimated or not (2 levels).

The sources of population process error that were used in the OMs or assumed in the EMs were on recruitment only (R), recruitment and changes in cohort abundance over time (R+S), or recruitment and natural mortality (R+M). We did not use the log-normal bias-correction feature for process errors or observations described by @stockmiller21 for operating and EMs [@lietal_inreview]. Simulations were all carried out on the University of Massachusetts Green High-Performance Computing Cluster. Code for completing the simulations and summarizing results can be found at https://github.com/timjmiller/SSRTWG/ecov_study/mortality.

@@

In the WHAM model, environmental covariates are assumed to be described as state-space processes with annual observations of the true latent covariate [@milleretal16;stockmiller21]. In our simulations, the latent covariate is assumed to be a stationary first order autoregressive (AR1) process
$$
X_y|X_{y-1} \sim \text{N}\left(\mu_E\left(1-\rho_E\right) + \rho_E E_{y-1}, \left(1-\rho_E^2\right)\sigma^2_E\right)
$$
with marginal mean $\mu_E=0$ and variance $\sigma^2_E$. The four configuration of the latent environmental covariate in the operating models assume the one of two values for the marginal standard deviation ($\sigma_E \in \{0.1, 0.5\}$) and for the autocorrelation parameter ($\rho_E \in \{0, 0.5\}$).

The observations of the latent environmental covariate are assumed to be unbiased and Gaussian
$$
x_y|X_y \sim \text{N}\left(X_y,\sigma^2_e\right)
$$
The standard deviation of the environmental observations in the operating models is one of two values $\sigma_E \in \{0.1, 0.5\}$. Figure \ref{om_ecov_example} provides example simulations of the latent process and observations under the alternative configurations.

@@

Many of the characteristics of the population biology and structure are the same as those in @milleretal_inreview1. We configures the stock with 10 age classes (ages 1 to 10+) where the last is a "plus" group that accumulates any older individuals. We model the population for 40 years with annual catch and index observations. We assume spawning occurs annually 1/4 of the way through the year. The maturity at age was a logistic function of age with age at 50\% maturity ($a_{50}$) = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top left). We assumed a von Beralanffy growth function for weight at age  with length at age defined as 
$$
L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
$$
where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a length-weight 
$$
W_a = \theta_1 L_a^{\theta_2}
$$
where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$ (Figure \ref{om_inputs_fig}, top right).

The general model for natural mortality in year $y$ is a log-linear function of a mean(median) parameter $\beta_{M}$, effects of an environmental covariate $E_y$ and a possibly autocorrelated normally distributed annual process error $\varepsilon_{M,y}$
$$
\log M_y = \beta_M + \beta_{E} E_y + \varepsilon_{M,y}
$$
$\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$ [@stockmiller21]. We assume the median natural mortality rate $e^{\beta_M} = 0.2$ is constant across ages. For R and R+S OMs and EMs, the process errors are set equal to 0. For all R+M OMs, we assume the same standard deviation $\sigma_M = 0.3$ and is estimated in the R+M EMs. The covariate effect is one of 3 alternative values in the operating models ($\beta_\text{E} \in \{0,0.25,0.5\}$. The parameters defining the simulated covariate time series, size of the covariate effect, and any natural mortality random effects result in a range of different levels of variation in annual natural mortality rates (Figure \ref{M_example}).

We assume is a single fishing fleet harvesting the population with a logistic selectivity function with age at 50\% selection $a_{50} = 5$ and slope = 1 (Figure \ref{om_inputs_fig}, bottom left).

We assumed expected recruitment each year from a Beverton-Holt stock-recruit relationship (SRR)
\[
R_{y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}}.
\]
All biological inputs to calculations of spawning biomass per recruit (i.e., weight, maturity, and natural mortality at age) are constant in the R and R+S OMs without covariate effects on natural mortality. Therefore, steepness and unfished recruitment are also constant over the time period for those OMs [@millerbrooks21] and with our assumed biological inputs and selectivity the fishing mortality resulting in 40\% reduction in spawning biomass per recruit is $F_{40\%} = 0.348$. With an assumed unfished recruitment of $R_0 = e^{10}$, setting $\Fmsy = \Fspr[40]$ results in a steepness of 0.69 and $a=0.60$ and $b = 2.4 \times 10^{-5}$ (Figure \ref{om_inputs_fig}, bottom right).

We used two fishing scenarios for OMs. In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years followed by fishing at \Fmsy for the last 20 years (denoted $2.5\Fmsy \rightarrow \Fmsy$). In the second scenario, the stock is fished at \Fmsy for the entire time period (40 years). The magnitude of the overfishing assumptions is based on average estimates of overfishing for NEUS groundfish stocks from @wiedenmannetal19 and similar to the approach in @legaultetal23.

We specified initial population abundance at age at the equilibrium distribution that corresponds to fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$. This implies that, for a deterministic model, the abundance at age would not change from the first year to the next.

For R+M OMs and all OMs with environmental covariate effects on natural mortality, steepness is not constant, but we used the same $a$ and $b$ parameters as other operating models which equates to a steepness and $R_0$ at the median of the time series process for M. For operating models with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same fishing history as other operating models which corresponds to \Fmsy at the mean selectivity parameters.

We configured all R, R+S, and R+M OMs with uncorrelated random effects on recruitment with standard deviation on log(recruitment) $\sigma_R = 0.5$. This same assumption was used by @milleretal_inreview1 for R+M OMs and other OMs with fishery selectivity and index catchability process errors. For R+S OMs, cohort temporal transition process errors were uncorrelated  with $\sigma_{2+} = 0.3$

### Catch and index observations {-}

We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1 (Figure \ref{om_inputs_fig}). This selectivity was used to define \Fmsy, \Fspr[40], and steepness for the Beverton-Holt stock recruitment parameters above. We assumed a logistic-normal distribution for the age-composition observations associated with the fleet where errors in the multivariate normal transformation are independent. The standard deviation parameter is also constant across ages.

@@

@@

\begin{itemize}
\item whether the mean natural mortality $\beta_M$ was estimated or assumed known ($\log 0.2$)
\item whether an environmental effect $\beta_E$ was estimated or not (fixed at 0)
\item whether the process errors were assumed on recruitment only (R), recruitment and survival (R+S), or recruitment and natural mortality (R+M)
\end{itemize}
The configuration of the process errors in the estimating models generally matched the corresponding options in the operating models.

@@

## Performance measures {-}

@@

We calculated median errors of
\begin{itemize}
\item $\beta_E$, the effect of environmental covariates on natural mortality, 
\item $\beta_M$, the mean log-natural mortality rate,
\end{itemize}
and the median relative errors of 
\begin{itemize}
\item annual natural mortality rate,
\item annual spawning stock biomass, and
\item annual fully-selected fishing mortality rate.
\end{itemize}

@@

For natural mortality rate, SSB, and fully-selected fishing mortality rate, we summarized results for each of the annual values, but we present results only for estimates from the first year (start), year 21 (middle), and the last year (end), because there were no appreciable differences between the results for these three years and those for the other years.

@@

Because @milleretal_inreview1 found inferences are most reliable for scenarios with low observation error for indices and age composition data and with temporal contrast in fishing pressure, we restrict our attention to scenarios with these characteristics in the main paper (temporal contrast in fishing and low observation error), but include corresponding results for the other scenarios in the Supplementary Materials.

## Convergence {-}

When the median natural mortality rate was assumed known, corresponding to usual practice in application of assessment models, good convergence for all EMs was observed for R+S operating models with \Fmsy fishing history, more variation in the latent environmental covariate ($\sigma_E = 0.5$), and lower error in the associated observations $\sigma_e = 0.1$ (Figure \ref{convergence}). Good convergence for all EMs was also observed for R+S OMs with the step change in fishing history, but with low error in indices and age comp observations.

EMs that assumed random effects just on recruitment (R), nearly always converged across all operating models that assumed more variation in the latent environmental covariate ($\sigma_E = 0.5$) and lower error in the associated observations $\sigma_E = 0.1$. EMs that assumed recruitment and survival random effects (R+S), had poor converge probability when the OMs had alternative process errors (R or R+M). The R+S EMs showed best convergence for the R+S OMs where there was lower error in the environmental observations $\sigma_E = 0.1$ or higher error in environmental observations of latent true environmental covariates with greater temporal variation. EMs that assumed recruitment and natural mortality random effects (R+M), had poor convergence probability for all OMs that had a step change in fishing history and higher uncertainty in indices and age composition.

@@

As might be expected, there was an overall drop in the probability of convergence when the mean natural mortality rate was estimated rather than assumed at the true value (Figure \ref{convergence}). Otherwise, general trends described above with mean natural mortality fixed, apply when estimated.

@@

When fitting the EMs to R and R+S OMs, AIC was generally accurate in determining the correct source of process errors and, when no effect of the covariate was simulated in the OMs, AIC was also accurate for the correct treatment of the environmental covariate $\beta_E = 0$ (Figs. \ref{aic}, \ref{aic_Rom}, \ref{aic_RSom}). However, when OMs included an effect of the covariate AIC accuracy for estimating an effect in the EM was only found when the OMs also included lower covariate observation error and higher effect size and higher temporal variability in the covariate. On the other hand, for R+M Oms, AIC generally was not able to accurately determine the correct source of process errors, but was able to accurately determine the correct treatment of the environmental covariate (Figs. \ref{aic} and \ref{aic_RMom}).

When estimating models assumed the mean natural mortality rate was known, the best accuracy of AIC for model selection occurred for models with R+S process errors. R+S estimating models ranked best with very low frequency for R or R+M operating models and with very high frequency for R+S operating models (Figure \ref{aic}). R estimating models were determined best with high frequency for R operating models, but also for R+M operating models. R+M estimating models were rarely determined best for any operating models including those where the process errors matched.

@@

Relative to the assumption that the mean natural mortality rate was known, estimating the mean natural mortality rate had small effects on the accuracy of AIC in selecting the appropriate process error and whether the covariate affect natural mortality (Figure \ref{aic}). Where there were differences there were small decreases in accuracy for determining the appropriate process error and determining a covariate effect when there was one.

@@

When the EMs assumed the mean natural mortality rate was known, we observed generally accurate estimation of environmental effects across all EM and OM process error assumptions and all true covariate effect sizes, when ther was low uncertainty in environmental observations and larger temporal contrast in the simulated true environmental covariate (Figure \ref{Ecov_beta_bias_M_fixed}). We observed a negative trend in bias of the environmental effect with increased effect size when temporal variation in the covariate was lower and/or uncertainty in the covariate observations was higher. When the OMs had R+S process errors with low temporal variation in the true environmental covariate and lower uncertainty in the indices age age composition, estimated covariate effects were highly variable.  In most cases the relative error of $\beta_E$ did not depend on the source of process error assumed in the EM. When there was an effect of the EM process error assumption it was when OMs had R+S process errors. The worst bias was observed when OMs assumed R+S process errors, high uncertainty in covariate observations, low variability in the covariate, and low uncertainty in index and age composition observations.

@@

We present results for error in annual natural mortality rate conditional on three alternative EM configurations for the natural mortality parameters: 1) mean natural mortality rate parameter is fixed at the true value ($\beta_M = log(0.2)$) and no covariate effect is assumed ($\beta_E = 0$),  2) $\beta_M = log(0.2)$ and $\beta_E$ is estimated, and 3) both $\beta_M$ and $\beta_E$ are estimated.

When OMs and EMs assume $\beta_M = log(0.2)$ and $\beta_E = 0$, there is no annual variation in natural mortality for OMs with R or R+S process errors simulated or assumed in the EM and, therefore, estimation bias is not possible (Figure \ref{annual_M_bias_M_fixed_beta_fixed}). We also observe little or no evidence of bias (confidence intervals include 0) for R or R+S EMs for any OMs even when the OMs included an effect of the covariate on natural mortality ($\beta_E > 0$). Including process error on M (R+M) produces more variability in errors for R and R+S EMs than including increased level of effect of the covariate on M. For R and R+S OMs, errors in annual M for R+S EMs were less variable than those for R EMs.

@@

Poor quality of biological, fishery, and observation may adversely affect ability to distinguish between true process and observational error [@puntetal14; @stewartmonnahan17; @croninpunt21; @fischetal23; @lietal_inreview_a].
